MSB1015_Assignment 2

### 2019-10-13 MSB1015_Assignment2_SuzannetenHage.Rmd
section 5: Changed validation to LOO, improved documentation
Section 5: 
* Fixed a small error that occurred because I was using an old name for the data (data.raw is changed into data_wikidata). 
* I realized it is better to use leave-one-out validation than cross-validation, because our dataset is small. So I changed this and added it to the documentation. 
* Added an instruction video on PLS to the documentation

The rJava package requires Java to be installed. The code has been developed using Java version 1.8.0_191. A tutorial on how to install this Java Version can be found here: https://downlinko.com/download-install-jdk-8-windows.html

@@

#Installation of missing packages 
if (length(setdiff(packages, rownames(installed.packages()))) > 0) {
    install.packages(setdiff(packages, rownames(installed.packages()))) 
    }

#Load packages
library("WikidataQueryServiceR")
library('rJava')
library('rcdk')
library('caTools')
library('pls')
library('Metrics')

@@

**2. Get the descriptors of the alkane structures**
In this section, I use four descriptors for all alkanes from CDK. I can obtain the descriptors based on the alkane smiles. These descriptors will be used later as predictors in the PLS model.

@@

#At the time of writing, this returns a dataframe of 23 variables describing 134 alkanes. 
```

**3. Data pre-processing**
Before we can use the descriptors and boiling points to train a PLS model, some data pre-processing needs to be done.

@@

for (i in 1:length(celcius)) {
   #Convert the value to kelvin
   data_wikidata$bp[celcius[i]] <- data_wikidata$bp[celcius[i]] + 273.15
   #Convert the label to kelvin. 
   data_wikidata$bpUnitLabel[celcius[i]] <- 'kelvin'
  }

@@

for (i in 1:length(fahrenheit)) {
   #Convert the value to kelvin
   data_wikidata$bp[fahrenheit[i]] <- (data_wikidata$bp[fahrenheit[i]] + 459.67) * (5/9)
   #Convert the label to kelvin. 
   data_wikidata$bpUnitLabel[fahrenheit[i]] <- 'kelvin'
  }

@@

**3b)** Combine the descriptors and boiling points in one dataset, so we can use this dataset for the PLS. I remove the columns with (almost) all zeros, because these descriptors do not have much predictive value.

@@

#At the time of writing, this returns a dataframe of 7 variables describing 134 alkanes. 
```

**3c)** To be able to test how well the model I train performs, it is necessary to keep a part of my data aside as testing data.

@@

#At the time of writing 134 alkanes are split into a training set of 115 alkanes and test set of 19 alkanes. 
```

@@

#Create null_model
null_model <- mean(train$data_wikidata.bp)

#Get error of null model
error_null_model <- rmse(test$data_wikidata.bp, null_model)

#Print the error
print(paste0("The error of the Null model is ", error_null_model))

@@

**5. Partial Least Squares**
Here, I train the PLS model with the training data, and use the model to predict the boiling points of the test data.

@@

#Training of the PLS model. 
PLS_model <- plsr(data_raw.bp ~ ., 
                  data = train, 
                  validation = "CV"
                  )

#test.try <- test[,c(1:34)];
#Predict the boiling points of the test set. 
predicted.bp <- predict(PLS_model, test)

#Calculate the error. 
error <- RMSEP(PLS_model, newdata = test)

@@

#Plot the predicted data vs. the original data. 
plot(test$data_raw.bp, predicted.bp[,,3], xlim=c(300,1000), ylim=c(300,1000))

@@

>>>>

The rJava package requires Java to be installed. The code has been developed using Java version 1.8.0_191. A tutorial on how to install this Java Version on windows can be found here: https://downlinko.com/download-install-jdk-8-windows.html

@@

#Installation of missing packages.
if (length(setdiff(packages, rownames(installed.packages()))) > 0) {
    install.packages(setdiff(packages, rownames(installed.packages()))) 
    }

#Load packages.
library("WikidataQueryServiceR")
library('rJava')
library('rcdk')
library('caTools')
library('pls')
library('Metrics')

@@

**2. Get the descriptors of the alkane structures**
In this section, I use four descriptors for all alkanes from CDK. I obtain the descriptors based on the alkane smiles. These descriptors are used later as predictors in the PLS model.

@@

#At the time of writing, this returns a dataframe of 23 variables describing 134 alkanes.

**3. Data pre-processing**
Before we use the descriptors and boiling points to train a PLS model, some data pre-processing needs to be done.

@@

for (i in 1:length(celcius)) {
   #Convert the value to kelvin.
   data_wikidata$bp[celcius[i]] <- data_wikidata$bp[celcius[i]] + 273.15
   #Convert the label to kelvin. 
   data_wikidata$bpUnitLabel[celcius[i]] <- 'kelvin'
  }

@@

for (i in 1:length(fahrenheit)) {
   #Convert the value to kelvin.
   data_wikidata$bp[fahrenheit[i]] <- (data_wikidata$bp[fahrenheit[i]] + 459.67) * (5/9)
   #Convert the label to kelvin. 
   data_wikidata$bpUnitLabel[fahrenheit[i]] <- 'kelvin'
  }

@@

**3b)** Next, I combine the descriptors and boiling points in one dataset, so we can use this dataset for the PLS. I remove the columns with (almost) all zeros, because these descriptors do not have much predictive value.

@@

#At the time of writing, this returns a dataframe of 7 variables describing 134 alkanes.

**3c)** To be able to test how well the model I train performs, it is necessary to keep a part of my data aside as test set. Therefore, I split the data into a training and test set. I use 25% of my dataset as test set.

@@

#At the time of writing 134 alkanes are split into a training set of 115 alkanes and test set of 19 alkanes.

@@

#Create null_model.
null_model <- mean(train$data_wikidata.bp)

#Calculate the error of the null model.
error_null_model <- rmse(test$data_wikidata.bp, null_model)

#Print the error.
print(paste0("The error of the Null model is ", error_null_model))

@@

**5. Partial Least Squares**
Here, I train the PLS model with the training data, and use the model to predict the boiling points of the test data. If you are unfamiliar with PLS, this is video a good introduction: https://www.youtube.com/watch?v=AxmqUKYeD-U. In the model I use leave-one-out (LOO) validation, because my data set is small.

@@

#Training of the PLS model, using leave-one-out validation.
PLS_model <- plsr(data_wikidata.bp ~ ., 
                  data = train, 
                  validation = "LOO"
                  )

#Calculate the error of the PLS model.
error_PLS <- RMSEP(PLS_model, newdata = test)

#Use the model to predict the boiling points of the test set.
predicted_bp <- predict(PLS_model, test)

@@

#Plot the predicted data vs. the original data. 
plot(test$data_raw.bp, predicted_bp[,,3], xlim=c(300,1000), ylim=c(300,1000))

@@

#Print the error
print(paste0("The error of the PLS model is ", error_PLS))
